Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-10-16 isolation_by_coalescence.tex
last big edit?

\begin{abstract}
A number of methods commonly used in landscape genetics
use analogies to electrical resistance on a network
to describe and fit barriers to movement across the landscape
using genetic distance data.
These are motivated by a mathematical equivalence between electrical resistance between two nodes
and the ``commute time'',
which is the mean time for a random walk on that network to leave one node, visit the other, and return.
Here, we describe the differences between commute time
and coalescence time, which more accurately models genetic distances,
and explore the consequences for inference.
We implement a Bayesian method to infer effective movement rates and population sizes 
under both these models,
and find that inference using commute times
can dramatically mislead in the presence of biased gene flow.
Forwards-time simulation with continuous geography demonstrates that
coalescence-based inference remains more accurate than resistance-based methods on realistic data,
and highlights the need for methods that do not approximate continuous space
by a discrete set of populations.
\end{abstract}

@@

Genetic relatedness is determined by past gene flow,
a product of individual or gamete movements across geographic space.
Genomes therefore retain the traces of this movement,
and can contribute to the inference of how a species moves across a landscape,
which is important for understanding how diseases spread, how species adapt,
and how to retain genetic diversity in threatened species.
This is a potentially important source of information,
as direct observation can be difficult \citep[especially of long-distance movement]{dispersal_estimation},
or even impossible, if some of the populations in question no longer exist.

The idea of ``resistance distance'' is an important tool in the lanscape genomics toolbox.
Introduced by \citet{mcrae2006isolation},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{resistance}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain \citep{levin}.
This measure therefore averages over all possible paths through the network between the two.
For this reason, 
if assigning local conductances based on the values of some landscape variable
results in a significant positive correlation between 
genetic distances between samples from different parts of the landscape
and the effective resistances between those locations,
then this is taken as good evidence that the landscape variable is a good indicator of
where gene flow occurs \citep{mcrae2007circuit,cushman2006complex,more_examples}.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is higher in flatter areas,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more readily through flatter areas,
perhaps along river valleys.
More recent methods \citep{petkova2016visualizing,hanks2013circuit} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

@@

\citet{mcrae2006isolation} showed that coalescence time and resistance distance are equal
(up to shift and scaling) in isotropic landscapes such as a ring \citep{matsen_wakeley},
and found that the two were highly correlated in several test cases.
However, all these test cases involved symmetric rates of gene flow,
which is likely not the case in many situations, e.g.,
in rivers \citep{morrissey2009maintenance,sundqvist2016directional,hanks2017modeling},
source-sink dynamics \citep{source_sink},
or after population expansions \citep{expansions}.

\plr{CHANGE TO mostly RESISTANCE STARTING AT SIMPLE EXAMPLE}

In this paper, we contrast 
coalescence time and resistance distance,
and develop an inference method for movement rates on a landscape
using both as estimated from genetic distances.
As we will see,
although the two methods are conceptually similar,
commute time inference may get the wrong answer if data derive from a coalescent process,
as does real genetic data, especially in the presence of asymmetry.
We also explore the question of identifiability.
Resistance-based methods are often used to predict distances based on a few layers of geographical data
(land cover, slope, etcetera).
Some methods find the combination of landscape layers to best explain genetic distance,
which allows very fine geographic resolution \citep{infer_resistance}.
Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
however, the geographical resolution is much coarser.
One reason for this difference in resolution
is that EEMS uses more computationally intensive Bayesian methods.
However, we show that this is a more general problem --
inference of movement rates, in either a coalescence or commute time framework,
is an ill-conditioned problem,
implying the need for regularization to obtain reliable inference.
This explains the results of \citet{graves2013current},
who found a large area of resistance parameters
that produced equivalently good fits to genetic distance
(i.e., nonidentifiability, analogous to a flat likelihood surface).

Talk about MAPS.

% \plr{I didn't find myself wanting to explain isolation by distance here:
%     I don't think we need to? Do you agree?}

@@

There are a number of ways to calculate measures of dissimilarity using genetic data \citep{distances}.
In this paper we use mean pairwise divergence:
for two genomes, it is calculated as the density of sites at which the two differ.
It is more common in the literature to use $F_{ST}/(1-F_{ST})$,
but the two are highly correlated for good theoretical reasons \citet{slatkin_fst,rousset1997genetic},
especially for SNP data.
The genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
from the two genomes back to their most recent common ancestor at that site;
under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
pairwise divergence is therefore an unbiased estimate of the mean time to most recent common ancestor,
averaged across the genome \citep{hudson2007variance,ralphXX}.
It is natural to model this process by following the two lineages back through time
(and across space)
until they find their most recent common ancestor,
i.e., until they coalesce.
It is common to model the geographic distribution of a species
as a collection of randomly mating populations
that exchange occasional migrants,
thus discretizing geographic space.
If the genome is selectively netural,
then the motion of these lineages back through time
forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
In this framework, each lineage performs a random walk across the populations,
with movement probabilities depending on the flux of individuals between populations.
This provides the link between genetic distance and random walks.

@@

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the jump rate from $x$ to $y$.
This means that if the Markov chain is at location $X_t$ at time $t$,
then for $x \neq y$, the probability the chain jumps from $x$ to $y$ is proportional to $G_{xy}$,
i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain --
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$.

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
Therefore, $H$ uniquely determines $G$.

% \paragraph{Commute times}
The \emph{commute times} are a symmetrized version of the hitting times
(which need not be symmetric):
\begin{align} \label{eqn:R_from_H}
\text{\bf (commute time)} \qquad
    R_{xy} = H_{xy} + H_{yx},
\end{align}
or $R = H + H^T$.
We use $R$ to refer to the commute time as a reminder that commute times are also effetive resistances,
and to reserve $C$ for the coalescence time.

Although commute times are uniquely determined by movement rates, the reverse is not true.
Commute times only depend on the symmetric part of $H$ --
given an skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$
(assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
(It is always true that rows of $G_\epsilon$ sum to zero.)
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{sylvester_eqn}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZC$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

\paragraph{Constraints and uniqueness}
We have seen that to find the movement rates of a continuous-time Markov chain
given the coalescence or commute times
we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{eqn:G_from_H} for $G$,
and that these do not have unique solutions.
The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
However, usually in applications the locations lie across geographical space,
so that many of the entries of $G$ can be assumed to be zero.
We can get an idea of this by simply counting equations and unknowns.
For concreteness, suppose that the spatial locations
are arranged in a square grid of $n$ locations,
so that each is connected to four others.
(Boundary locations will have fewer, but we omit this detail.)
Since movement rates in each direction can be different,
there are $4n$ free parameters, each corresponding to an offdiagonal entry of $G$.
Coalescence rates provide another $n$ parameters.

Since commute times are symmetric, and $R_{zz} = 0$, 
equation \eqref{eqn:R_from_H} only provides $n (n-1)/2$ informative equations.
This is larger than $4n$, the number of parameters,
for a grid with at least nine nodes (i.e., $3 \times 3$).
The same calculation holds for commute times, 
except that (a) we get an additional $n$ equations from the diagonal entries,
but (b) this is counteracted by the additional $n$ coalescence parameters.

This suggests that although nonuniqueness of solutions may not be a problem in practice,
as long as geography is discretized into sufficiently many locations.
However, another problem appears at finer geographic resolutions:
although solutions may be in principle unique,
the problems become \emph{ill-conditioned} 
in the sense that arbitrarily small variations in the input times
(even numerical instability)
may produce large differences in the inferred rates.
Very roughly speaking, this occurs because as the spatial resolution increases,
the matrix $G$ converges to a second-order elliptic differential operator \citep{pde};
such operators are deformations of the Laplacian \citep{laplacian},
which is well-known to have eigenvalues that decay exponentially \citep{laplacian_eigenvals}.
Information about the landscape contained in higher-order modes
(e.g., finer resolution changes in hitting times)
is then almost obliterated by the small eigenvalues of $G$,
making the inverse problem possible in theory but not in practice.
For more discussion of related problems, see \citet{badtruth,shape_of_a_pop,inverse_prob}.

\paragraph{When are coalescence and commute times equal?}
If in a given biological situation the two times were equal --
i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
then inference with either model would be equivalent.
We have already shown that this is not the case in general,
but it turns out that it \emph{is} true
under some restrictive assumptions not uncommonly found in population genetics.
We show in Appendix \ref{sec:com_eq_coal} that
if hitting times are symmetric ($H = H^T$),
then this does occur,
and so the two methods are equivalent for symmetric island models
(where the populations are arranged on a ring 
and migration rates only depend on the distance between them).
Hitting times are not symmetric for a square grid with uniform migration
(the mean time to hit the center from a corner is less than the reverse),
but the square grid is quite close to the torus, which does have symmetric hitting times.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{The simplest example}

Consider a Markov chain with two states, 
where the rate of movement from state 1 to state 2 is $G_{12}$
and the rate of movement from state 2 to state 1 is $G_{21}$,
so the generator matrix is
\begin{align*}
G = 
    \begin{bmatrix}
        -G_{12}  & G_{12} \\
         G_{21}  & -G_{21}
    \end{bmatrix}.
\end{align*}
Equation \ref{eqn:C_matrix} equates two $2 \times 2$ matrices,
so provides four equations. Only three of these are unique;
simplifying these and using that $C_{12} = C_{21}$, these are:
\begin{align*}
    2 \left(C_{12} - C_{11} \right) G_{12} - C_{11} \gamma_1 &= -1 \\
    \left(C_{11} - C_{12} \right) G_{21} + \left(C_{22} - C_{12}\right) G_{12} &= -1 \\
    2 \left( C_{12} - C_{22} \right) G_{21} - C_{22} \gamma_2 &= -1 .
\end{align*}
Given $C$, we have three equations for the four unknowns, 
$G_{12}$, $G_{21}$, $\gamma_{1}$, and $\gamma_{2}$,
which we can solve symbolically.
The valid solutions to these equations are those with $G_{12}$, $G_{21}$, $\gamma_1$,
and $\gamma_2$ nonnegative.
If $C_{11} \neq C_{12}$, 
we can write the solution with $\gamma_1$ as the free variable:
\begin{align*}
G_{12} &= \frac{1}{2(C_{11} - C_{12})} - \frac{C_{11}}{2(C_{11} - C_{12})}\gamma_1 \\
G_{21} &= \frac{-2(C_{11} - C_{12}) + (C_{12} - C_{22})}{2(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})}{2(C_{11} - C_{12})^2}\gamma_1 \\
\gamma_2 &= \frac{(C_{11} - 2C_{12} + C_{22})^2}{C_{22}(C_{11} - C_{12})^2}
	- \frac{C_{11}(C_{12} - C_{22})^2}{C_{22}(C_{11} - C_{12})^2}\gamma_1.
\end{align*}
This implies that $\gamma_2$ decreases as $\gamma_1$ increases as long as $C_{12} \neq C_{22}$.
This makes sense because
increasing a rate of coalescence cannot make expected coalescence times longer,
so in order to keep coalescence times the same,
if one coalescence rate is increased, another must be decreased.
This also implies that, in order for $\gamma_2$ to be a value other than $0$, 
we must have $C_{11} - 2C_{12} + C_{22} \neq 0$.
In general, if there is a value or range of values of $\gamma_1$
for which the other movement and coalescence parameters are non-negative,
then a solution exists.

@@

Inference with commute times is done in exactly the same way,
except that $q$ replaces $\gamma$,
and instead of $\mathcal{C}(G,\gamma)$ we use $\comdistfn(G, q)$,
which is the matrix $\comdist$ computed from $G$ and $q$ by solving equation \eqref{eq:commute_approx}.

@@

\paragraph{MCMC methods}
To sample from the posterior distribution of equation \eqref{eq:post}, 
we used R \citep{R} 
to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
Starting locations were chosen by sampling from the prior distribution.
Before the usual ``burn-in'' phase before sampling for inference begins,
we included a period of ``pre-burn-in'' which used the same MCMC procedure
with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
portion of parameter space.
Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
and $3 \times 10^6$ iterations of burn-in,
followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
although for small graphs fewer iterations were necessary.
We used standard methods to assess convergence and mixing.

@@

These simulations are expected to be nearly equivalent to population-based simulations
(either forwards-time or coalescent);
the only difference being that noise terms about the theoretical mean should be somewhat correlated.

\paragraph{Forwards-time simulations}
To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017slim,haller2018treesequences},
with individuals living across continuous, two-dimensional geography (sometimes with barriers),
from which we recorded genomic data.
The basic simulation, which we modified to produce several other situations,
is as follows.
Each simulation had around $10^4$ diploid, hermaphroditic individuals, 
living across two-dimensional geography.
Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
truncated at $3 \sigma_d$.
Simulations were done on either a square region with width equal to $80\sigma_d$,
or a $3:5$ rectangular region with the same area.
Offspring are dispersed away from the mother using the same kernel,
reflected from the boundaries of the habitable area.
In each simulation, $\sigma_d$ was chosen 
so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
we also model local competition. 
To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
between each pair of nearby individuals;
then, the fitness of each individual is set to $\exp(-x/40)$, 
where $x$ is the sum of their interaction strengths with other individuals.
This reduces fitness for individuals in denser areas.
Individuals are chosen to reproduce proportional to their fitness.
Clumpiness is similar to what is seen in 
Figure \ref{fig:5x3b_continuous}.

Individuals have 10 pairs of ``chromosomes'' with $2 \times 10^5$ loci each 
and a recombination rate of $10^{-8}$ per generation per locus.
Neutral mutations were added at a rate of $10^{-9}$ per meiosis.
To make these simulations computationally feasible,
the mutation rate was set to 0.0 during the forwards simulation,
while tree sequences were recorded \citep{haller2018treesequences},
and mutations were added after the fact with msprime,
which is equivalent to including mutations during the forwards simulation 
but resulted in much faster run times \citep{kelleher2018xxx}.
In all cases, the number of mutations was more than adequate to reliably estimate genetic distance.

>>>>

\begin{abstract}
A number of methods commonly used in landscape genetics
use an analogy to electrical resistance on a network
to describe and fit barriers to movement across the landscape
using genetic distance data.
These are motivated by a mathematical equivalence between electrical resistance 
between two nodes of a network
and the ``commute time'',
which is the mean time for a random walk on that network to leave one node, visit the other, and return.
Here, we describe the differences between commute time
and coalescence time, which more accurately models genetic distances,
and explore the consequences for inference.
We implement a Bayesian method to infer effective movement rates and population sizes 
under both these models,
and find that inference using commute times
can dramatically mislead in the presence of biased gene flow.
We then use forwards-time simulation with continuous geography to demonstrate that
coalescence-based inference remains more accurate than resistance-based methods on realistic data,
but difficulties highlight the need for methods 
that explicitly model continuous, heterogeneous geography.
\end{abstract}

@@

Genetic relatedness is determined by past gene flow,
a product of individual or gamete movements across geographic space.
Genomes therefore retain the traces of this movement,
and can contribute to the inference of how a species moves across a landscape,
which is important for understanding how diseases spread, how species adapt,
and how to retain genetic diversity in threatened species.
This is a potentially important source of information,
as direct observation can be difficult 
(especially of long-distance movement, \citep{cayuelademographic,levin_et_al_seed_dispersal}),
or even impossible, if some of the populations in question no longer exist.

The idea of ``resistance distance'' is an important tool in the landscape genomics toolbox.
Introduced by \citet{mcrae2006isolation},
it makes use of a mathematical equivalence between random walks and electrical resistance
\citep{nashwilliams1959random}:
the expected time for a continuous-time Markov chain that starts at node $x$
to first hit node $y$ and then return to $x$ (the ``commute time'')
is equal to the effective resistance between $x$ and $y$ in an electrical network
whose conductances are given by the movement rates of the Markov chain \citep{levin2008markov,doyle2006random}.
This measure therefore averages over all possible paths through the network between the two.
For this reason, 
if assigning local conductances based on the values of some landscape variable
results in a significant positive correlation between 
genetic distances between samples from different parts of the landscape
and the effective resistances between those locations,
then this is taken as good evidence that the landscape variable is a good indicator of
where gene flow occurs \citep{mcrae2007circuit,cushman2006complex}.
For instance,
if the conductance across a grid cell in a discretization of the landscape
is higher in flatter areas,
and this produces effective resistance values 
that are significantly correlated with genetic distances,
then one might conclude that the species in question tends to disperse more readily through flatter areas,
perhaps along river valleys.
More recent methods \citep{petkova2016visualizing,hanks2013circuit} seek to build \emph{de novo} 
a map of conductance values that produces resistances 
most strongly correlated with observed genetic distance,
and then interpret regions of low conductance in the resulting map as barriers to gene flow.

@@

\citet{mcrae2006isolation} showed that coalescence time and resistance distance are equal
(up to shift and scaling) in isotropic landscapes such as a ring \citep{matsen2006convergence},
and found that the two were highly correlated in several test cases.
However, all these test cases involved symmetric rates of gene flow,
which is likely not the case in many situations, e.g.,
in rivers \citep{morrissey2009maintenance,sundqvist2016directional,hanks2017modeling},
source-sink dynamics \citep{dias1996sources},
or after population expansions \citep{with2002landscape}.

In this paper, we contrast 
coalescence time and resistance distance,
and develop an inference method for movement rates on a discrete landscape
using both as estimated from genetic distances.
As we will see,
although the two methods are conceptually similar,
commute time inference may get the wrong answer if data derive from a coalescent process,
especially in the presence of asymmetry.
We also explore the question of identifiability.
Resistance-based methods are often used to predict distances based on a few layers of geographical data
(land cover, slope, etcetera).
Some methods find the combination of landscape layers to best explain genetic distance,
which allows very fine geographic resolution \citep{shaffer2017desert}.
Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
however, the geographical resolution is much coarser.
A recently-released method 
that is based on EEMS but uses haplotype length data and an underlying model of coalescence time
\citep{alasadi2018estimating}
is promising, but still produces similarly coarse maps.
One reason for this difference in resolution
is that EEMS uses more computationally intensive Bayesian methods.
However, we show that this is a more general problem,
both because of unmodeled ``process noise'' involved in discretization of geographical populations,
and the mathematical structure of the underlying problem.

@@

There are a number of ways to calculate measures of dissimilarity using genetic data.
In this paper we use mean pairwise divergence:
for two genomes, it is calculated as the density of sites at which the two differ.
It is more common in the literature to use $F_{ST}/(1-F_{ST})$,
but the two are highly correlated for good theoretical reasons 
\citet{slatkin1991inbreeding,rousset1997genetic},
especially for SNP data.
The two genomes differ at a site only if there has been a mutation in some ancestor on the lineages leading
from the two genomes back to their most recent common ancestor at that site;
under an infinite-sites model with average mutation rate $\mu$ per site and per meiosis,
pairwise divergence divided by $\mu$
is therefore an unbiased estimate of the mean time to most recent common ancestor,
averaged across the genome \citep{hudson2007variance,ralph2015empirical}.
It is natural to model this process by following the two lineages back through time
(and across space)
until they find their most recent common ancestor,
i.e., until they coalesce.
Suppose we model the geographic distribution of the species
as a collection of randomly mating populations
that exchange occasional migrants,
thus discretizing geographic space.
If the genome is selectively neutral,
then the motion of these lineages back through time
forms a Markov process, the \emph{structured coalescent} \citep{wakeley2005coalescent}.
In this framework, each lineage performs a random walk across the populations,
with movement probabilities depending on the flux of individuals between populations.
This provides the link between genetic distance and random walks.

@@

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the jump rate from $x$ to $y$.
This means that if the Markov chain is at location $X_t$ at time $t$,
then for $x \neq y$, the probability the chain jumps from $x$ to $y$ in the next unit of time
is proportional to $G_{xy}$,
i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain --
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$.

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous-fill-2014}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
Therefore, $H$ uniquely determines $G$.

% \paragraph{Commute times}
The \emph{commute times} are a symmetrized version of the hitting times
(which need not be symmetric),
defining $R = H + H^T$, i.e.:
\begin{align} \label{eqn:R_from_H}
\text{\bf (commute time)} \qquad
    R_{xy} = H_{xy} + H_{yx}.
\end{align}

Although commute times are uniquely determined by movement rates, the reverse is not true.
Commute times only depend on the symmetric part of $H$ --
given an skew-symmetric matrix $Z$ (so that $Z + Z^T = 0$),
any Markov chain that has hitting times given by $H_\epsilon = H + \epsilon Z$ for some $\epsilon$
will have the same commute times.
The resulting hitting times may not be valid --
if we let $G_\epsilon$ be the matrix constructed by applying equation \eqref{eqn:G_from_H} to $H_\epsilon$,
% (assuming this is invertible),
then it is not guaranteed that the offdiagonal elements of $G_\epsilon$ are all nonnegative, as required.
(It is always true that rows of $G_\epsilon$ sum to zero.)
We provide a small concrete example in Appendix \ref{ss:simple_example}.
However, since $G$ and $H$ are continuous functions of each other,
if all offdiagonal elements of $G$ are strictly positive,
then there exists a positive $\epsilon_0$ such that all $G_\epsilon$ for $\epsilon < \epsilon_0$
do define valid Markov chains, all with the same commute times.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{bhatia1997solve}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZC$ is skew-symmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $\epsilon$
so that $G + \epsilon Z$ remains the generator of a valid Markov chain.

\paragraph{Constraints and uniqueness}
We have seen that to find the movement rates of a continuous-time Markov chain
given the coalescence or commute times
we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{eqn:G_from_H} for $G$,
and that these do not have unique solutions.
The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
However, usually in applications the locations lie across geographical space,
so that many of the entries of $G$ can be assumed to be zero.
This reduces the number of unknowns to solve for,
and so can render the solution unique.
We can get an idea of this by simply counting equations and unknowns.
For concreteness, suppose that the spatial locations
are arranged in a square grid of $n$ locations,
so that each is connected to four others.
(Boundary locations will have fewer, but we omit this detail.)
Since movement rates in each direction can be different,
there are $4n$ free parameters, each corresponding to an off-diagonal entry of $G$.
Coalescence rates provide another $n$ parameters.

Since commute times are symmetric, and $R_{zz} = 0$, 
equation \eqref{eqn:R_from_H} provides $n (n-1)/2$ informative equations.
This is larger than $4n$, the number of parameters,
for a grid with at least nine nodes (i.e., $3 \times 3$),
so the system of equations will usually have a unique solution
for grids larger than this, except in special cases.
The same calculation holds for commute times, 
except that (a) we get an additional $n$ equations from the diagonal entries,
but (b) this is counteracted by the additional $n$ coalescence parameters.

This suggests that although nonuniqueness of solutions may not be a problem in practice,
as long as geography is discretized into sufficiently many locations.
However, another problem appears at finer geographic resolutions:
although solutions may be in principle unique,
the problems become \emph{ill-conditioned} 
in the sense that arbitrarily small variations in the input times
(even numerical instability)
may produce large differences in the inferred rates.
Roughly speaking, this occurs because as the spatial resolution increases,
the matrix $G$ converges to a second-order elliptic differential operator \citep{stroock1997multidimensional};
such operators are deformations of the Laplacian \citep{feller4},
which is well-known to have eigenvalues that decay rapidly 
\citep{hpmckean1967curvature,kuttler1984eigenvalues}.
Information about the landscape contained in higher-order modes
(e.g., finer resolution changes in hitting times)
is then almost obliterated by the small eigenvalues of $G$,
making the inverse problem possible in theory but not in practice.
For more discussion of related problems, see 
\citet{epstein2008badtruth,myers2008learn,terhorst2015fundamental}.

\paragraph{When are coalescence and commute times equal?}
If in a given biological situation commute and coalescence times are equal --
i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
then inference with either model would be equivalent.
This was demonstrated in some situations by \citet{mcrae2006isolation}.
We have already shown that this is not the case in general,
but it turns out that it \emph{is} true
under some restrictive assumptions not uncommonly found in abstract population genetics models.
We show in Appendix \ref{sec:com_eq_coal} that
if hitting times are symmetric ($H = H^T$),
then this does occur,
and so the two methods are equivalent for symmetric island models
(where the populations are arranged on a ring 
and migration rates only depend on the distance between them).
Hitting times are not symmetric for a square grid with uniform migration
(the mean time to hit the center from a corner is less than the reverse),
but the square grid is quite close to the torus, which does have symmetric hitting times.

@@

Inference with resistance distance is done in exactly the same way,
except that $q$ replaces $\gamma$,
and instead of $\mathcal{C}(G,\gamma)$ we use $\comdistfn(G, q)$,
which is the matrix $\comdist$ computed from $G$ and $q$ by solving equation \eqref{eq:commute_approx}.

@@

\paragraph{MCMC methods}
To sample from the posterior distribution of equation \eqref{eq:post}, 
we used R \citep{Rmanual} 
to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
Starting locations were chosen by sampling from the prior distribution.
Before the usual ``burn-in'' phase before sampling for inference begins,
we included a period of ``pre-burn-in'' which used the same MCMC procedure
with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
portion of parameter space.
Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
and $3 \times 10^6$ iterations of burn-in,
followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
although for small graphs fewer iterations were necessary.
We used standard methods to assess convergence and mixing.

@@

These simulations are expected to be nearly equivalent to neutral, discrete population-based simulations
(either forwards-time or coalescent);
the only difference being that noise terms about the theoretical mean should be somewhat correlated.

\paragraph{Forwards-time simulations}
To produce realistic data, we implemented forwards-time simulations using SLiM v3.1 
\citep{haller2018forward,haller2018treesequence},
with individuals living across continuous, two-dimensional geography (sometimes with barriers),
from which we recorded genomic data.
The basic simulation, which we modified to produce several other situations,
is as follows.
Each simulation had around $10^4$ diploid, hermaphroditic individuals, 
living across two-dimensional geography.
Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
truncated at $3 \sigma_d$.
Simulations were done on either a square region with width equal to $80\sigma_d$,
or a $5:3$ rectangular region with the same area.
Offspring are dispersed away from the mother using the same kernel,
reflected from the boundaries of the habitable area.
In each simulation, $\sigma_d$ was chosen 
so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
we also model local competition. 
To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
between each pair of nearby individuals;
then, the fitness of each individual is set to $\exp(-x/40)$, 
where $x$ is the sum of their interaction strengths with other individuals.
This reduces fitness for individuals in denser areas.
Individuals are chosen to reproduce proportional to their fitness.
Clumpiness is similar to what is seen in 
Figure \ref{fig:5x3b_continuous}.

Individuals have 10 pairs of ``chromosomes'' with $2 \times 10^5$ loci each 
and a recombination rate of $10^{-8}$ per generation per locus.
Neutral mutations were added at a rate of $10^{-9}$ per meiosis.
To make these simulations computationally feasible,
the mutation rate was set to 0.0 during the forwards simulation,
while tree sequences were recorded \citep{haller2018treesequence},
and mutations were added after the fact with msprime,
which is equivalent to including mutations during the forwards simulation 
but resulted in much faster run times \citep{kelleher2018efficient}.
